Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-11-04 analysis/paper/paper.Rmd
SPECIES ORDER ADJUSTED;MINOR EDITS

The global climate disruption is causing considerable reduction in
Arctic sea ice extent, volume, and seasonal presence [@kwok2018,
@meier2014, @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the
region [@]. Such disruptions are a particular cause
of concern for ice-associated seals (ringed (*Pusa hispida*), bearded
(*Erignathus barbatus*), spotted (*Phoca largha*), and ribbon seals
(*Histriophoca fasciata*)) that depend on spring and early summer sea
ice (March-June) in the Bering and Chukchi seas as a platform for
important functions, such as pupping, nursing, breeding, and molting
[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
and large knowledge gaps complicate predictions about the ultimate
effects of changes in sea ice on the behavior, abundance, and
distribution of these seals. Knowledge about evolutionary constraints on
the timing of reproductive and molting behavior is generally lacking, so
it is difficult to predict the readiness with which ice-associated seal
species will be able to adapt to future changes (e.g., by adjusting
pupping or molting schedules to earlier dates or different locales).
Additionally, trends in abundance of these species are largely unknown,
so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

@@

In this study, we used data from 15 years of bio-logger records in a
comprehensive investigation of haul-out behavior by 3 species of
ice-associated seals in the Bering and Chukchi seas. We focused our
effort on bearded, spotted, and ribbon seals and did not include ringed
seals in the analysis. Ringed seals are unique in their use of snow
lairs in the spring months, and they warrant a separate study that
includes additional environmental covariates, such as the onset of snow
melt and snow depth. Our goals are threefold. First, we wish to
establish baseline estimates for the chronology of haul-out behavior in
the critical spring season for each species across different age and sex
classes. Second, we wish to refine estimates of haul-out availability
corrections for aerial surveys in order to improve estimates of seal
abundance. Previously estimated availability correction factors (e.g.,
[@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not
investigate meteorological variables that have been shown to influence
haul-out behavior of walruses [@reder2003; @udevitz2009]. Third,
we explored the annual variability in haul-out timing and possible
linkage to dramatic changes in the availability and extent of seasonal
sea ice between 2005 and 2020. Our work extends the scope of previous
haul-out analyses, includes the influence of meteorological variability,
and investigates the potential impact of a changing icescape on the
behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, spotted, and ribbon seals was filtered
to include only records from 1 March to 15 July between
`r min(lubridate::year(analysis_data$haulout_dt))` and
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of
the 'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers
(Redmond, Washington, USA) and either adhered to the hair on the seal or
attached to the rear flipper inter-digital webbing. The use of
bio-loggers adhered to the back or head provide some benefits over
flipper mounted devices (e.g. increased satellite transmittal rates,
locations at sea) but these are lost during the annual molt, which
limits the effective length of haul-out timelines. Additionally, sensors
attached to the head or dorsal region are often dry while the seal is at
the surface and can slightly bias the percent-dry values recorded by the
device. For this study, in cases where both tag types were deployed,
percent dry observations from the flipper tag were preferred. Sex and
age class (non-dependent *young-of-the-year*, sexually immature
*subadults*, and mature *adults*) were determined at the time of
deployment by subjective judgement based on length, and pelage
characteristics for some species. Field identification of age class can
be inexact, particularly when discerning subadults from adults. In the
case of ribbon seals, pelage differences between subadults and adults
are distinguishable, with subadults often having less defined ribbons.
Bearded seal subadults will often have a spotted pattern in the pelage
that is not seen in adults. Spotted seal pelage is monomorphic and can
not be used to discern age class. Despite these challenges, we feel the
age classifications used in this analysis are useful categories in
testing for age-related effects on haul-out behavior. Seals determined
to be less than one year were classified as young-of-the-year. Subadults
are those seals likely greater than one year of age but not yet sexually
mature. Adults are likely sexually mature and older than approximately
four years. For those tags deployed on young-of-the-year and
transmitting into the next year, the age class was advanced to subadult
on 1 March of the following year. Table \@ref(tab:tblDeploy) provides a
summary of these deployments and data received from them.

@@

Tags that fall off due to molt or mortality and remain on ice or land
can still send data to satellites, and we did not want to include data
from these tags in our analyses. As such, end times of each deployment
were identified by experts who examined several simultaneous data
streams (e.g., tag locations and dive behavior) to determine when tags
ceased providing data consistent with seal behavior. Sensor readings
outside of these start and end times were discarded prior to analysis.
After approximately 9 months, 7 devices deployed on the rear flipper of
bearded seals reported implausible sensor data. All data after the first
instance of unrealistic values were censored from this analysis. Figure
\@ref(fig:dataCal) shows a calendar distribution of these data across
the study season for each species. Observations are concentrated in the
months of May and June due to the timing of deployment (April and May)
and the timing of molt (May and June). During molt, seals (and their
attached bio-loggers) spend more time out of the water and more data is
transmitted. Molt timing also impacts when many deployments end as any
tags adhered to the hair will fall off.

```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.3), breaks = c(100,500,1000,1500)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  scale_y_discrete(limits = rev) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        #axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

@@

Models for all species include the following effects: day-of-year, solar
hour, temperature, wind speed, barometric pressure, precipitation, wind
chill (represented by a *wind:temperature* interaction
[@udevitz2009]), and day-of-year and time of day interactions to
permit diurnal patterns to change throughout the year. Spotted and
ribbon seal models included age-sex class and interactions between
day-of-year and age-sex class, but we omitted these from bearded seal
models due to poor representation of age-sex classes (Table
\@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a
substantial range and we were interested in possible differences in the
timing of haul-out behavior along a latitudinal gradient. We omitted the
latitudinal effect from ribbon and spotted seal models since, during the
spring, these species are most prevalent near the southern ice edge in
the Bering Sea [@conn2014a].

@@

@@

Visualizing the marginal effect of each weather covariate on haul-out
probability is difficult in this analysis because of the collinearity
between covariates as well as the temporal variation in covariate values
during the study period. The relationship of each weather covariate with
haul-out probability, averaged over the other weather conditions, is
more variable than simply showing model coefficients would imply. Figure
\@ref(fig:ribbonHOwx)) presents the predicted haul-out probability of
ribbon seals across the range of weather conditions encountered in the
observed data. The collinearity and temporal variation are apparent in
the visual patterns that emerge. By

@@

Age and sex class were not included in the model for bearded seals and,
as such, results are shown for all ages. Unlike ribbon and spotted
seals, bearded seals were predicted to have a bi-modal distribution of
haul-out probability across the day (Figure \@ref(fig:beardedHOCal)). In
addition to a peak around local solar noon, the bearded seal model
predicts additional haul-out activity around local midnight. The
haul-out behavior is also more protracted throughout the spring season
compared to ribbon and spotted seals.

Like ribbon and spotted seals, bearded seal haul-out probability appears
most influenced by wind
($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
= `r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`)
and temperature
($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$
= `r bearded_type3$temp2$F.value`; $p$ =
`r bearded_type3$temp2$Prob.F`). Additionally, wind chill was also a
significant influence
($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$
= `r bearded_type3$temp2_wind$F.value`; $p$ =
`r bearded_type3$temp2_wind$Prob.F`). Any influence of barometric
pressure or precipitation is less apparent (Figure
\@ref(fig:beardedHOwx)).

```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data. Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability. Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}

@@

p <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(yday, ho_prob)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.25, 
                  stroke = 0, color = "grey80") +
  geom_pointrange(data = peak_ho_df,
                  mapping = aes(yday, ho_prob, ymin = 0, ymax = ho_prob, color = year),
                  shape = 15) +
  scale_color_brewer(palette = "Paired") +
  scale_x_continuous(breaks = c(74,105,135,166),
                     labels = c("15-Mar", "15-Apr", "15-May", "15-Jun")) +
  facet_grid(age_sex~species) +
  theme(legend.position = "bottom") +
  xlab("") + ylab("haul-out probability") +
  ggtitle("Annual variability in the timing of peak haul-out probability")
 p
```

@@

In this paper, we modeled data from bio-logging sensors deployed on
bearded, spotted, and ribbon seals to examine factors affecting haul-out
behavior on sea ice in the Bering and Chukchi seas. Our analysis shows
seals use sea ice progressively more through the spring, with peak use
near mid-May to early June before declining again. This pattern aligns
well with what has been previously documented and confirms our
bio-logging deployments are likely capturing population-level behavioral
patterns. Seals preferentially haul out on ice shortly after solar noon,
which allows seals to maximize absorption of solar radiation
[@feltz1966]. Interestingly, bearded seals appear to have two peaks in
haul-out activity across the day, one shortly after solar noon, and one
centered near solar midnight. A similar bi-modal pattern has been seen
in ringed seals [@vonduyke2020] and suggests that bearded and ringed
seals may be operating under different constraints than ribbon and
spotted seals. Bearded and ringed seals are distributed across higher
latitudes and the extended daylight hours may allow more flexibility in
optimizing resting periods with foraging.

Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class
on haul-out probabilities for all species except bearded seals because
of low sample size. While both ribbon and spotted seals exhibited a
unimodal diel haul-out pattern generally centered around local solar
noon, there were key differences across species, age, and sex that match
expectations given what we know of their ecological behavior. Spotted
seals are known to form triads during the breeding season [@burns2009,
@mcclintock2015] where a female and dependent pup are accompanied on the
ice by a suitor male. The male waits for the female to wean the pup and
enter estrus, and fends off any other males that may be interested.
Triad formation results in both males and females spending a large
portion of the day hauled out on ice and a protracted spring haul-out
season for both sexes. Because mating occurs in the water, females may
also be less inclined to venture out for foraging trips while still
nursing the pup. We see this reflected in the predicted haul-out
behavior, with both males and females exhibiting a broad distribution of
time out of the water throughout the solar day and the season. Ribbon
seals are not known to form triads and our model predicts a progression
of increased haul-out behavior with females starting earlier in the
season than males. Notably, female ribbon seals spend a large portion of
the day in the water during the pupping period, aligning with the
hypothesis that ribbon seal females continue foraging while nursing. In
the case of both ribbon and spotted seals, subadults are the first to
begin consistent haul-out behavior and follow a typical phocid pattern
where subadults molt first as they do not have any reproductive
constraints.

@@

Our model uncovered annual deviations in the timing and magnitude of
haul-out peaks for ribbon and spotted seals. The timing of peak haul-out
activity appears to fall within a relatively narrow time window of 3-4
weeks in late May and early June. This remarkable level of consistency
across 15 years is indicative of the adaptations these species have
evolved to synchronize their key life history stages with favorable
environmental conditions (e.g. availability of sea ice, increased
oceanographic productivity, favorable weather conditions).
Unfortunately, this may also be indicative of a limited range of
plasticity as climatological conditions change. Additionally, while from
an ecological perspective the haul-out behavior appears consistent, the
interannual differences in timing and magnitude are large enough to have
important ramifications on calculations of abundance and trend.

>>>>

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus barbatus*),
ribbon seals (*Histriophoca fasciata*), and spotted (*Phoca largha*)) that
depend on spring and early summer sea ice (March-June) in the Bering and Chukchi
seas as a platform for important functions, such as pupping, nursing, breeding,
and molting [@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
and large knowledge gaps complicate predictions about the ultimate effects of
changes in sea ice on the behavior, abundance, and distribution of these seals.
Knowledge about evolutionary constraints on the timing of reproductive and
molting behavior is generally lacking, so it is difficult to predict the
readiness with which ice-associated seal species will be able to adapt to future
changes (e.g., by adjusting pupping or molting schedules to earlier dates or
different locales). Additionally, trends in abundance of these species are
largely unknown, so it is difficult to say what effect, if any, declines in sea
ice habitat have had, or will have, on seal densities.

@@

In this study, we used data from 16 years of bio-logger records in a
comprehensive investigation of haul-out behavior by 3 species of ice-associated
seals in the Bering and Chukchi seas. We focused our effort on bearded, ribbon,
and spotted seals and did not include ringed seals in the analysis. Ringed seals
are unique in their use of snow lairs in the spring months, and they warrant a
separate study that includes additional environmental covariates, such as the
onset of snow melt and snow depth. Our goals are threefold. First, we wish to
establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes.
Second, we wish to refine estimates of haul-out availability corrections for
aerial surveys in order to improve estimates of seal abundance. Previously
estimated availability correction factors (e.g., [@bengtson2005a], [@conn2014a],
and [@verhoef2014a]) accounted for variables such as the time-of-day and
day-of-year, but did not investigate meteorological variables that have been
shown to influence haul-out behavior of walruses [@reder2003; @udevitz2009].
Third, we explored the annual variability in haul-out timing and possible
linkage to dramatic changes in the availability and extent of seasonal sea ice
between 2005 and 2020. Our work extends the scope of previous haul-out analyses,
includes the influence of meteorological variability, and investigates the
potential impact of a changing icescape on the behavior of these species.

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, ribbon, and spotted seals was filtered to
include only records from 1 March to 15 July between `r min(lubridate::year(analysis_data$haulout_dt))` 
and `r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
annual molt, which limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while
the seal is at the surface and can slightly bias the percent-dry values recorded
by the device. For this study, in cases where both tag types were deployed,
percent dry observations from the flipper tag were preferred. Sex and age class
(non-dependent *young-of-the-year*, sexually immature *subadults*, and mature
*adults*) were determined at the time of deployment by subjective judgement
based on length, and pelage characteristics for some species. Field
identification of age class can be inexact, particularly when discerning
subadults from adults. In the case of ribbon seals, pelage differences between
subadults and adults are distinguishable, with subadults often having less
defined ribbons. Bearded seal subadults will often have a spotted pattern in the
pelage that is not seen in adults. Spotted seal pelage is monomorphic and can
not be used to discern age class. Despite these challenges, we feel the age
classifications used in this analysis are useful categories in testing for
age-related effects on haul-out behavior. Seals determined to be less than one
year were classified as young-of-the-year. Subadults are those seals likely
greater than one year of age but not yet sexually mature. Adults are likely
sexually mature and older than approximately four years. For those tags deployed
on young-of-the-year and transmitting into the next year, the age class was
advanced to subadult on 1 March of the following year. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fall off due to molt or mortality and remain on ice or land can still
send data to satellites, and we did not want to include data from these tags in
our analyses. As such, end times of each deployment were identified by experts
who examined several simultaneous data streams (e.g., tag locations and dive
behavior) to determine when tags ceased providing data consistent with seal
behavior. Sensor readings outside of these start and end times were discarded
prior to analysis. After approximately 9 months, 7 devices deployed on the rear
flipper of bearded seals reported implausible sensor data. All data after the
first instance of unrealistic values were censored from this analysis. Figure
\@ref(fig:dataCal) shows a calendar-like distribution of these data across the
study season for each species. Observations for ribbon and spotted seals are
concentrated in the months of May and June due to the timing of deployment
(April and May) and the timing of molt (May and June). During molt, seals (and
their attached bio-loggers) spend more time out of the water and more data is
transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off. Bearded seal deployments are more evenly
spread with fewer observations in May and June. The majority of bearded seal
deployments start later in the summer and by May tags have either fallen off or
the batteries are depleted.

```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.3), breaks = c(100,500,1000,1500)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  scale_y_discrete(limits = rev) +
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        #axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("The number of haul-out behavior observations varies from March through July",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time of day interactions to permit diurnal patterns to change
throughout the year. Ribbon and spotted seal models included age-sex class and
interactions between day-of-year and age-sex class, but we omitted these from
bearded seal models due to poor representation of age-sex classes (Table
\@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect (and an
interaction with day-of-year), since bearded seals occupy a substantial range
and we were interested in possible differences in the timing of haul-out
behavior along a latitudinal gradient. We omitted the latitudinal effect from
ribbon and spotted seal models since, during the spring, these species are most
prevalent near the southern ice edge in the Bering Sea [@conn2014a].

@@

Age and sex class were not included in the model for bearded seals and, as such,
results are shown for all ages. Bearded seals were predicted to have a bi-modal
distribution of haul-out probability across the day (Figure
\@ref(fig:beardedHOCal)). In addition to a peak around local solar noon, the
bearded seal model predicts additional haul-out activity around local midnight.
The haul-out behavior is also more protracted throughout the spring season
compared to ribbon and spotted seals.

Like ribbon and spotted seals, bearded seal haul-out probability appears
most influenced by wind
($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
= `r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`)
and temperature
($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$
= `r bearded_type3$temp2$F.value`; $p$ =
`r bearded_type3$temp2$Prob.F`). Additionally, wind chill was also a
significant influence
($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$
= `r bearded_type3$temp2_wind$F.value`; $p$ =
`r bearded_type3$temp2_wind$Prob.F`). Any influence of barometric
pressure or precipitation is less apparent.

```{r beardedHOwx,fig.height = 5.5, include=FALSE, fig.id = "beardedHOwx", fig.cap="Variability in predicted haul-out probability of bearded seals across the range of weather conditions encountered in the observed data. Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability. Solid lines are smoothed fits to the predictions and provided to better indicate the marginal effect."}

@@

Visualizing the marginal effect of each weather covariate on haul-out
probability is difficult in this analysis because of the collinearity
between covariates as well as the temporal variation in covariate values
during the study period. The relationship of each weather covariate with
haul-out probability, averaged over the other weather conditions, is
more variable than simply showing model coefficients would imply. Figure
\@ref(fig:ribbonHOwx) presents the predicted haul-out probability of
ribbon seals across the range of weather conditions encountered in the
observed data. The collinearity and temporal variation are apparent in
the visual patterns that emerge.

@@

@@

pal <- c(RColorBrewer::brewer.pal(12, "Paired"), "black")

p <- ggplot(plot_df %>% filter(age_sex != "YOUNG OF YEAR"), aes(yday, ho_prob)) +
  geom_pointrange(aes(ymin = lower95, ymax = upper95), alpha = 0.25, 
                  stroke = 0, color = "grey80") +
  geom_pointrange(data = peak_ho_df,
                  mapping = aes(yday, ho_prob, ymin = 0, ymax = ho_prob, color = year),
                  shape = 15) +
  scale_color_discrete(type = pal) +
  scale_x_continuous(breaks = c(74,105,135,166),
                     labels = c("15-Mar", "15-Apr", "15-May", "15-Jun")) +
  facet_grid(age_sex~species) +
  theme(legend.position = "bottom") +
  xlab("") + ylab("haul-out probability") +
  ggtitle("Annual variability in the timing of peak haul-out probability")
 p
```

@@

In this paper, we modeled data from bio-logging sensors deployed on bearded,
ribbon, and spotted seals to examine factors affecting haul-out behavior on sea
ice in the Bering and Chukchi seas. Our analysis shows seals use sea ice
progressively more through the spring, with peak use near mid-May to early June
before declining again. This pattern aligns well with what has been previously
documented and confirms our bio-logging deployments are likely capturing
population-level behavioral patterns. Seals preferentially haul out on ice
shortly after solar noon, which allows seals to maximize absorption of solar
radiation [@feltz1966]. Interestingly, bearded seals appear to have two peaks in
haul-out activity across the day, one shortly after solar noon, and one centered
near solar midnight. A similar bi-modal pattern has been seen in ringed seals
[@vonduyke2020] and suggests that bearded and ringed seals may be operating
under different constraints than ribbon and spotted seals. Bearded and ringed
seals are distributed across higher latitudes and the extended daylight hours
may allow more flexibility in optimizing resting periods with foraging. Other
factors such as polar bear predation (which is rare for ribbon and spotted
seals in the Bering Sea) may also explain differing haul-out patterns.

Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
While both ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match expectations given what we know of their
ecological behavior. Ribbon seals are not known to form triads and our model
predicts a progression of increased haul-out behavior with females starting
earlier in the season than males. Notably, female ribbon seals spend a large
portion of the day in the water during the pupping period, aligning with the
hypothesis that ribbon seal females continue foraging while nursing. Spotted
seals are known to form triads during the breeding season [@burns2009,
@mcclintock2015] where a female and dependent pup are accompanied on the ice by
a suitor male. The male waits for the female to wean the pup and enter estrus,
and fends off any other males that may be interested. Triad formation results in
both males and females spending a large portion of the day hauled out on ice and
a protracted spring haul-out season for both sexes. Because mating occurs in the
water, females may also be less inclined to venture out for foraging trips while
still nursing the pup. We see this reflected in the predicted haul-out behavior,
with both males and females exhibiting a broad distribution of time out of the
water throughout the solar day and the season.  In the case of both ribbon and
spotted seals, subadults are the first to begin consistent haul-out behavior and
follow a typical phocid pattern where subadults molt first as they do not have
any reproductive constraints.

@@

Our model uncovered annual deviations in the timing and magnitude of haul-out
peaks for ribbon and spotted seals. The timing of peak haul-out activity appears
to fall within a relatively narrow time window of 3-4 weeks in late May and
early June. This consistency across 15 years is indicative of the adaptations
these species have evolved to synchronize their key life history stages with
favorable environmental conditions (e.g. availability of sea ice, increased
oceanographic productivity, favorable weather conditions). Unfortunately, this
may also be indicative of a limited range of plasticity as climatological
conditions change. Additionally, while from an ecological perspective the
haul-out behavior appears consistent, the interannual differences in timing and
magnitude are large enough to have important ramifications on calculations of
abundance and trend.
